Building Markov State Models for Periodically Driven Non-Equilibrium Systems

### 2015-01-26 writedown/period.noneq.msm/submit.jctc.reply/paper.tex
add source of the paper

\begin{abstract}
Recent years have seen an increased interest in non-equilibrium molecular dynamics (NEMD) simulations, especially for molecular systems with periodic forcing by external fields, e.g., in the context of studying  effects of electromagnetic radiation on the human body tissue. Lately, an NEMD methods with local thermostating has been proposed that allows for studying 
non-equilibrium processes in a statistically reliable and thermodynamically consistent way. In this article, we demonstrate how to construct Markov State Models (MSMs) for such NEMD simulations. MSM building has been well-established for systems in equilibrium where MSMs with just a few (macro-)states allow for accurate reproduction of the essential kinetics of the molecular system under consideration. Non-equilibrium MSMs have been lacking so far. The article presents how to construct such MSMs and illustrates their validity and usefulness for the case of conformation dynamics of alanine dipeptide in an external electric field.  \\
\textbf{Keywords:} {Non-equilibrium, Markov states model, Alanine dipeptide, Electric field, Floquet theory}
\end{abstract}

@@

Biomolecular systems under non-equilibrium conditions caused by external fields, especially
systems under periodic forcing, have attracted increasing
interest recently.  For example, the potential effects of electromagnetic
radiation on the human body tissue (e.g.~DNA, protein, and membrane)  has been extensively
investigated in a vast number of articles, with the following list just representing an incomplete selection~\cite{bohr2000microwave, bohr2000microwave-1, dePomerai2000cell,
  dePomerai2003microwave, mancinelli2004non, inskip2001cellular, bekard2013electric, budi2005electric,
  budi2007effect, budi2008comparative, astrakas2012structural,
  damm2012can, english2009nonequilibrium, solomentsev2012effects}.  Molecular dynamics (MD) simulations have proved particularly useful for understanding
the response of biomolecular conformations to external fields because of their ability to  resolve molecular
details that sometimes cannot be resolved in experiments. Only
recently, a non-equilibrium MD simulation (D-NEMD) method with local thermostating has been proposed~\cite{wang2014exploring} that allows for studying 
non-equilibrium processes in a statistically reliable and thermodynamically consistent way. 
Despite the significance of the non-equilibrium phenomena, the
analysis of the non-equilibrium MD simulations mainly follows standard approaches, and reliable tools for quantitative description of
the essential conformational dynamics of the molecular system under external forcing are still lacking.

@@

Despite their many advantages, MD simulations have severe limitations. For example, one has to assume that the underlying force fields are appropriately describing the internal and external molecular interactions, and the maximal possible simulation length often is shorter than the timescale of interest. 
This article is mainly concerned with circumventing the latter obstacle by introducing non-equilibrium Markov State Models. Markov State Models
(MSM) have been well developed over the
past decade in theory~\cite{A19-31,prinz2011markov} and applications (see the recent book~\cite{A19-1} for an overview), and software implementations~\cite{A19-49, MSMBuilder}, but for systems under equilibrium conditions only!  
The principal idea of equilibrium MSMs is to approximate the original high-dimensional MD system by
a reduced Markovian dynamics over a finite number of (macro-)states. These (macro-)states
have to be identified with the dominant metastable sets, in the sense that
typical MD trajectories stay in the vicinity of a metastable set substantially longer than
the systems needs for a transition to another such state~\cite{A19-31,schuette2011markov}. In this case, the metastable sets are the main conformations of the molecular system under consideration which, often enough, are given by the main wells in the energy landscape. 
It has been shown that for molecular systems exhibiting such metastable sets the Markovian dynamics given by an MSM allows very close approximation of the longest relaxation processes of the underlying molecular system, at least under equilibrium conditions~\cite{sarich2010approximation,Eigenvalues}. 
Moreover, it has been demonstrated that in such cases MSM building requires short MD trajectories only, much shorter than the timescales of interest~\cite{PNAS09,kohlhoff2014cloud}.  
Thus, MSM building often allows to study the dynamical behavior on long timescales without requiring MD trajectories of comparable length.
Moreover, MSMs are utilized for understanding very long MD simulations: Extracting the essential structures and dynamical properties from long MD runs is
becoming increasingly difficult as the system size and trajectory length grow; MSMs have been used to construct kinetic fingerprints from MD simulations~\cite{A19-39} that help understanding the essential dynamics and in comparison with experiments~\cite{PrinzKellerNoe_PCCP11_Perspective}.

@@

% \textbf{Importance: first application of MSM in a non-equilibrium system.}
To the knowledge of the authors this work is the first attempt of using MSMs for analyzing non-equilibrium systems under periodic external forcing.
More precisely, we will demonstrate how to use MSMs for investigate the conformational dynamics of a peptide (alanine dipeptide) under an oscillating electric field (EF).
To this end, we will show how to generalize Markov state modeling to periodic non-equilibrium conditions where one cannot assume reversibility of the dynamics as it is mostly done in the literature on MSM building.

The outline of the article is as follows: 
In Sec.~\ref{sec:disc}, we discuss the temporal and spatial
discretizations needed to construct an MSM. We consider spatial discretization
of the dihedral angle space in the traditional sense of full partition MSMs~\cite{pande2010everything,A19-29}.
In the temporal direction, the non-equilibrium process
is discretized utilizing Floquet's theorem. This results in a time-homogeneous, 
but not necessarily reversible, that is, irreversible Markov process. Since  full spatial partition does not make any sense
for high-dimensional systems, we next show how to construct a \emph{few-state} MSM based on milestoning~\cite{schuette2011markov,A19-29} of the discretized irreversible Markov process in Sec.~\ref{sec:build-msm}.
The validity of the discretizations and the resulting MSM is checked in Sec.~\ref{sec:alanine}
by comparing the kinetic fingerprints given by the MSM to brute force
non-equilibrium MD simulations of alanine dipeptide under oscillating EF.
The findings are summarized in Sec.~\ref{sec:conclusion} including a chart presenting
the workflow of MSM building for non-equilibrium systems, and a list of open questions.

@@

\section{Non-equilibrium molecular dynamics and its discretization}
\label{sec:disc}
We consider diffusive molecular dynamics in an energy landscape $V$ driven by the time-dependent external driving force $E(t)D(x_t)$ with the
$T$-periodic external field $E(t)$:
\begin{align}
  \label{eq:disc-1}
  d\vect x_t = \Big(-\nabla V(\vect x_t) + E(t) D(\vect x_t)\Big)dt + \sqrt{2\beta^{-1}} d\vect w_t, 
\end{align}
where $x_t\in\Omega$ denotes the state of the molecular system at time $t$ in state space $\Omega$, $\vect w_t$ denotes standard $n$-dimensional Brownian motion, and $\beta$ the inverse temperature,
i.e.~$\beta = 1/(k_B\mathcal T)$.
Thermostatted Hamiltonain or Langevin dynamics can be treated in the same way as explained herein, so for sake of simplicity we focus on the discussion of diffusive dynamics.
The propagation of probability
densities $\rho=\rho(\vect x,t)$ based on this kind of dynamics in the sense
of $\rho(\vect x,t)dx=\prob[\vect x_t\in [\vect x,\vect x+d\vect x)]$ is governed by
Fokker-Planck equation:
\begin{align}
  \label{eq:disc-fp}
  \frac{\partial \rho}{\partial t} = \ml^\dagger(t) \rho,
\end{align}
where $\ml^\dagger(t)$ is the adjoint of the generator
\begin{align}
  \label{eq:disc-3}
  \ml(t)=\beta^{-1}\Delta_{\vect x}+\Big(-\nabla_x V(\vect x) + E(t)D(\vect x)\Big)\cdot\nabla_{\vect x},
\end{align}
where $\Delta_{\vect x}$ denotes the Laplacian operator and $\nabla_{\vect x}$
the nabla-operator wrt to $\vect x$. 
The periodicity of the external driving force induces the periodicity of the generator,
i.e.~$\ml(t) = \ml(t+T)$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Spatial discretization: Master equation}
We will now introduce an appropriate spatial discretization of this kind of non-equilibrium MD --  this is done for reasons of simplicity only; we could completely avoid it for the price of more technical arguments.
For achieving this discretization, we 
introduce a partition of state space $\Omega$ into a finite number of disjoint
sets $\{ \Omega_1, \cdots, \Omega_N\}$ satisfying $\Omega = \cup_i \Omega_i$,
$\Omega_j\cap \Omega_j = \emptyset,\ \forall i\neq j$.
Utilizing the procedure described in Ref.~\cite{latorre2011structure} the original Fokker-Planck equation~(\ref{eq:disc-fp})
is discretized, resulting in a time-inhomogeneous Markov jump process in state
space $S = \{1, \cdots, N\}$ with time-dependent rate
matrix $\vect L(t) \in \mathbb R^{N\times N}$ satisfying
\begin{align}\label{eq:disc-4}
\sum\limits_{j=1}^N L_{ij}(t) & =  0\\ \label{eq:disc-5}
L_{ij}(t) & \ge  0, \quad i\not= j\\
L_{ij}(t) & =  L_{ij}(t+T)
\end{align}
for all real time $t\geq 0$.
Moreover, the rate matrix $L$ has the form $\vect L(t)=\vect L_0+E(t)\vect L_1$
where $E(t)$ is periodic with period $T>0$.
In analogy to \eqref{eq:disc-fp}, the Markov jump process generated by
$\vect L(t)$ transports probability distributions according to the associated Master equation
\begin{align}
  \label{eq:disc-master}
  \frac{d\vect p(t)}{dt} = \vect L^{\top}(t)\cdot \vect p(t)
\end{align}
where $\vect L^{\top}(t)$ denotes the matrix transpose of $\vect L(t)$, $\vect p(t)$ is an $N$-vector denoting the probability distribution on $S$ at time $t$, $p(i,t)$, for example, the probability to be in state $i$ (which corresponds to set $\Omega_i$) at time $t$.
As usual the properties (\ref{eq:disc-4}) and (\ref{eq:disc-5}) of
$\vect L(t)$ guarantee that the total probability mass is conserved,
i.e., if $p(i,0)\ge 0$ componentwise, then $p(i,t)\ge 0$ and $\sum_i
p(i,t) = \sum_ip(i,0)$.
% The solution of the master equation need no be
% periodic.
The temporal evolution of the probability distribution $\vect p(t)$ can be formally written
\begin{align}  \label{eq:disc-8}
\vect p(t)=\myphi(t)\vect p(0)
\end{align}
by using the
associated propagator matrix $\myphi(t)\in\mathbb R^{N\times N}$ that solves
\begin{align}
  \label{eq:disc-master-phi}
  \frac{d}{dt}\myphi(t) = \vect L^{\top}(t)\myphi(t), \quad \myphi(0) = \id .
\end{align}
Since the last equation can be considered column-wise, the propagator matrix inherits column-wise conservation properties:
$\Phi_{ij}(t) \ge  0$
and $\sum\limits_{i=1}^N \Phi_{ij}(t)  =  1$,
that is, $\myphi^{\top}(t)$ is a
stochastic matrix satisfying $\myphi^{\top}(t)\vect e=\vect e$
with $\vect e=(1,\ldots,1)^{\top}\in \mathbb R^N$.
Regarding these considerations, we find
\begin{align}
\label{eq:disc-10}  
\myphi_{ij}(t)=\prob\left(\vect X_t=i\mid \vect X_0=j \right),
\end{align}
where $\vect X_t$ denotes the Markov process generated by $\vect L(t)$.

The discretization sets that we used to go from $\vect x_t$ and $\ml(t)$ to $\vect X_t$
and $\vect L(t)$, respectively, can be assumed to provide an arbitrarily fine
partition of the original state space; then the transport properties
of $\vect L(t)$ are almost perfect approximations of the transport properties
of $\ml(t)$, in particular the approximation $p(i,t)\approx \prob(x_t\in \Omega_i)$ is almost perfect.

@@

As an effect of the periodicity of $\vect L(t)$ the propagator $\myphi(t+T)$
satisfies
\begin{equation}\label{compo-1}
\myphi(t+T)=\myphi(t)\myphi(T),
\end{equation}
for all $t\ge 0$. This can be seen by considering $\vect Y(t)=\myphi(t+T)$. It satisfies
\[
\frac{d\ }{dt}\vect Y(t)=\vect L^{\top}(t+T) \vect Y(t)=\vect L^{\top}(t)\vect Y(t),\quad \vect Y(0)=\myphi(T).
\]
When we consider this identity column-wise and use the propagator property of $\myphi(t)$ we get $\myphi(t+T)=\vect Y(t)=\myphi(t)\myphi(T)$. As a consequence of (\ref{compo-1}) we get for all integers $m=0,1,2,\ldots$ that 
\begin{equation}\label{compo-2}
\myphi(t+mT)=\myphi(t)\myphi^m(T).
\end{equation}
In combination with Eq.~\eqref{eq:disc-8}, we therefore
know the solution $\vect p(t)$ of the Master equation for all $t\ge 0$,
if we can compute $\myphi(t)$ for $t\in (0,T)$.
This is known as the Floquet theorem~\cite{floquet1883equations}.
In particular we get the long-term evolution of the propagator:
\begin{align}
\label{eq:floq-13}  
\myphi(mT)=\myphi^m(T),
\end{align}
where $\myphi^m(T)$ denotes the $m$th power of $\myphi(T)$. Thus, for the probability at integral periods we have
\begin{align}
  \label{eq:floq-dynamics}
  p(mT) =  \myphi(mT)\, p(0) = \myphi^m(T)\, p(0).
\end{align}

Using the Floquet theorem, the time-inhomogeneous Markov jump process $X_t$
is therefore discretized into a \emph{time-homogeneous} (not necessarily
reversible) Markov jump process $\tilde X_{m} = X_{mT}, \ m\in\mathbb
N$, which is generated by transition matrix
\begin{equation}\label{P}
\vect P=\myphi^{\top}(T).
\end{equation}
We prefer to consider the discrete-time process $\tilde X_{m}$ instead of the time-continuous process $X_t$
because the powerful theories and
computational tools for time-homogeneous Markov processes can be directly applied. It worth noting that many of these tools require the transition matrix $P$ to satisfy the detailed balance condition. The computations in the Appendix~\ref{sec:app-revs} show that $P$ will in general \emph{not} satisfy this condition; in fact  the deviation {from} reversibility can be estimated from the work of the periodic driving does to the system.
There is no doubt that information within one period is lost by using this temporal
discretization, however, information regarding the long-term behavior of the system on timescales
much longer than the period will be perfectly described because of $\tilde X_{m} = X_{mT}\approx x_{mT}$ whenever our spatial discretization is fine enough.
At the same time,
the computational cost of generating $\tilde X_{m}$ is much less demanding
than the brute force simulations of NEMD, which implies lower
statistically uncertainty in calculating the observables of interest.

Since $P$ is a stochastic matrix, its eigenvalues are contained in the unit circle in
the complex plane, i.e., each eigenvalue $\lambda$ (potentially complex-valued)
satisfies $|\lambda|\le 1$. Furthermore $\lambda=1$ is an
eigenvalue with right eigenvector $\vect e=(1,\ldots,1)^{\top}$ and a left eigenvector $\mymu$
satisfying
$\mymu^{\top}P=\mymu^{\top}$.
From now on, we assume $P$ to be irreducible and aperiodic such that the Perron-Frobenius theorem holds, so the eigenvector corresponding to the eigenvalue $\lambda=1$ is  non-negative componentwise, and unique (up to normalization $\sum_i\mu(i)=1$). In this case $\mymu$ is the stationary measure in the sense that
$
 \mymu^{\top} P^m = \mymu^{\top},\ m\in\mathbb N
$,
% \begin{align}
% \label{eq:floq-14}  
% \end{align}
and (more precisely) the asymptotic evolution of an initial probability distribution $\vect p(t=0)$ by the process satisfies
$
\vect p^{\top}(0)\,P^m \to \mymu^{\top},\ m\to\infty,
$
% \begin{align}
% \label{eq:floq-15}  
% \end{align}
so that $\mymu$ can be seen as the quasi-stationary distribution of the non-stationary process.

% due to the discretization in both the spacial and temporal directions.
% In the numercal example in Sec.~\ref{sec:alanine}, we
% compute the

@@

% \subsection{Building the Markov State Model from the Floquet transition matrix}
If the discretization cells $\Omega_i$, $i=1,\ldots,N$ form a fine
partition of the molecular state space, the Markov chain defined via
the transition matrix $P$
 is discrete in time, but in space it still is a fine-scale description of the transport properties of the
dynamics with a very large number $N$ of states.  Now we want to coarse our description much further by
constructing a Markov State Model (MSM) for $\vect P$ with $K\ll N$
\emph{macrostates} that should be the metastable states of the system: The resulting $K\times K$ MSM transition matrix $\hat{\vect P}$
then defines the coarse grained long term kinetics that shall
approximate the original long term kinetics well. 
The idea behind MSM building is that given the molecular system under consideration exhibits metastable conformations then it is usually possible to construct a
relatively small number of discrete sets --the metastable sets that form the so-called macrostates-- that
correctly describe the slow  dynamics, and 
in each set the fast dynamics relaxes on some timescales significantly shorter than the metastable timescales.
Then if the MSM dynamics reproduces the slow timescales and the corresponding transitions
of the original dynamics~\eqref{eq:disc-1},
the former is considered to be a good approximation of the latter.

>>>>

\begin{abstract}
Recent years have seen an increased interest in non-equilibrium molecular dynamics (NEMD) simulations, especially for molecular systems with periodic forcing by external fields, e.g., in the context of studying the  effects of electromagnetic radiation on human body tissues. Lately, and NEMD method with local thermostating has been proposed; that allows the study of 
non-equilibrium processes in a statistically reliable and thermodynamically consistent way. In this article, we demonstrate how to construct Markov State Models (MSMs) for such NEMD simulations. MSM building is well-established for systems in equilibrium, where MSMs with only a few (macro-)states allow for accurate reproduction of the essential kinetics of the molecular system under consideration. Non-equilibrium MSMs have not been established yet. This article presents a method for construction of such MSMs and illustrates their validity and usefulness for the case of conformation dynamics of alanine dipeptide in an external electric field.  \\
\textbf{Keywords:} {Non-equilibrium, Markov states model, Alanine dipeptide, Electric field, Floquet theory}
\end{abstract}

@@

Biomolecular systems under non-equilibrium conditions caused by external fields, especially
systems under periodic forcing, have attracted increasing
interest recently.  For example, the potential effects of electromagnetic
radiation on  human body tissues (e.g.,~DNA, protein, and membrane)  have been extensively
investigated in a vast number of articles, with the following list  representing an incomplete selection~\cite{bohr2000microwave, bohr2000microwave-1, dePomerai2000cell,
  dePomerai2003microwave, mancinelli2004non, inskip2001cellular, bekard2013electric, budi2005electric,
  budi2007effect, budi2008comparative, astrakas2012structural,
  damm2012can, english2009nonequilibrium, solomentsev2012effects}.  Molecular dynamics (MD) simulations have proven particularly useful for understanding
the response of biomolecular conformations to external fields, because of their ability to  resolve molecular
details that sometimes cannot be resolved in experiments. Only
recently, a non-equilibrium MD simulation (D-NEMD) method with local thermostating has been proposed~\cite{wang2014exploring} that allows for studying 
non-equilibrium processes in a statistically reliable and thermodynamically consistent way. 
Despite the significance of the non-equilibrium phenomena, 
analysis of the non-equilibrium MD simulations mainly follows standard approaches, and reliable tools for quantitative description of
the essential conformational dynamics of the molecular system under external forcing remain unavailable.

@@

Despite their many advantages, MD simulations have severe limitations. For example, one has to assume that the underlying force fields  describe the internal and external molecular interactions appropriately, and the maximal possible simulation length often is shorter than the timescale of interest. 
This article is mainly concerned with circumventing the latter obstacle by introducing non-equilibrium Markov State Models. Markov State Models
(MSM) have been  developed thoroughly over the
past decade in theory~\cite{A19-31,prinz2011markov} and applications (see the recent book~\cite{A19-1} for an overview), and software implementations~\cite{A19-49, MSMBuilder}, but for systems under equilibrium conditions only.
The main concept underlying equilibrium MSMs is to approximate the original high-dimensional MD system 
using reduced Markovian dynamics over a finite number of (macro-)states. These (macro-)states
have to be identified with the dominant metastable sets, in the sense that
typical MD trajectories remain close to a metastable set substantially longer than
they need for a transition to another such state~\cite{A19-31,schuette2011markov}. In this case, the metastable sets are the main conformations of the molecular system under consideration, which, often enough, are given by the main wells in the energy landscape. 
It has been shown that for molecular systems exhibiting such metastable sets, the Markovian dynamics given by an MSM allows very close approximation of the longest relaxation processes of the underlying molecular system, at least under equilibrium conditions~\cite{sarich2010approximation,Eigenvalues}. 
Moreover, it has been demonstrated that, in such cases, MSM building requires short MD trajectories only, which are much shorter than the timescales of interest~\cite{PNAS09,kohlhoff2014cloud}.  
Thus, MSM building often allows the study of dynamic behavior on long timescales without requiring MD trajectories of comparable length.
Moreover, MSMs are utilized for understanding very long MD simulations. Extracting the essential structures and dynamic properties from long MD runs is
becoming increasingly difficult as the system size and trajectory length grow; MSMs have been used to construct kinetic fingerprints from MD simulations~\cite{A19-39} which facilitates understanding of essential dynamics and comparison with experimental data~\cite{PrinzKellerNoe_PCCP11_Perspective}.

@@

% \textbf{Importance: first application of MSM in a non-equilibrium system.}
To the knowledge of the authors this work is the first attempt of using MSMs for analyzing non-equilibrium systems under periodic external forcing.
More precisely, we will demonstrate how to use MSMs to investigate the conformational dynamics of a peptide (alanine dipeptide) under an oscillating electric field (EF).
To this end, we will show how to generalize Markov state modeling to periodic non-equilibrium conditions, where one cannot assume reversibility of the dynamics as is most often case in the literature on MSM building.

The outline of the article is as follows: 
In Section~\ref{sec:disc}, we discuss the temporal and spatial
discretizations needed to construct an MSM. We consider spatial discretization
of the dihedral angle space in the traditional sense of full partition MSMs~\cite{pande2010everything,A19-29}.
In the temporal direction, the non-equilibrium process
is discretized utilizing Floquet's theorem. This results in a time-homogeneous, 
but not necessarily reversible, that is, irreversible Markov process. Since  full spatial partition is not rational
for high-dimensional systems, we next show how to construct a \emph{few-states} MSM, based on milestoning~\cite{schuette2011markov,A19-29} of the discretized irreversible Markov process in Section~\ref{sec:build-msm}.
The validity of the discretizations and the resulting MSM is checked in Section~\ref{sec:alanine}
by comparing the kinetic fingerprints given by the MSM to brute force
non-equilibrium MD simulations of alanine dipeptide under oscillating EF.
The findings are summarized in Section~\ref{sec:conclusion}, which includes a chart presenting
the workflow of MSM building for non-equilibrium systems, and a list of open questions.

@@

\section{Non-equilibrium molecular dynamics and its discretization}
\label{sec:disc}
We consider diffusive molecular dynamics in an energy landscape $V$, driven by the time-dependent external driving force $E(t)D(x_t)$, with the
$T$-periodic external field $E(t)$:
\begin{align}
  \label{eq:disc-1}
  d\vect x_t = \Big(-\nabla V(\vect x_t) + E(t) D(\vect x_t)\Big)dt + \sqrt{2\beta^{-1}} d\vect w_t, 
\end{align}
where $x_t\in\Omega$ denotes the state of the molecular system at time $t$ in state space $\Omega$, $\vect w_t$ denotes standard $n$-dimensional Brownian motion, and $\beta$ the inverse temperature,
i.e.,~$\beta = 1/(k_B\mathcal T)$.
Thermostatted Hamiltonian or Langevin dynamics can be treated in the same way as explained herein, so for the sake of simplicity, we focus on the discussion of diffusive dynamics.
The propagation of probability
densities $\rho=\rho(\vect x,t)$ based on this kind of dynamics in the sense
of $\rho(\vect x,t)dx=\prob[\vect x_t\in [\vect x,\vect x+d\vect x)]$, is governed by
the Fokker-Planck equation:
\begin{align}
  \label{eq:disc-fp}
  \frac{\partial \rho}{\partial t} = \ml^\dagger(t) \rho,
\end{align}
where $\ml^\dagger(t)$ is the adjoint of the generator
\begin{align}
  \label{eq:disc-3}
  \ml(t)=\beta^{-1}\Delta_{\vect x}+\Big(-\nabla_x V(\vect x) + E(t)D(\vect x)\Big)\cdot\nabla_{\vect x},
\end{align}
where $\Delta_{\vect x}$ denotes the Laplacian operator and $\nabla_{\vect x}$
the nabla-operator with respect to to $\vect x$. 
The periodicity of the external driving force induces the periodicity of the generator,
i.e.,~$\ml(t) = \ml(t+T)$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Spatial discretization: Master equation}
We will now introduce an appropriate spatial discretization of this kind of non-equilibrium MD --  this is done for reasons of simplicity only; we could completely avoid it for the price of more technical arguments.
For achieving this discretization, we 
introduce a partition of state space $\Omega$ into a finite number of disjoint
sets $\{ \Omega_1, \cdots, \Omega_N\}$ satisfying $\Omega = \cup_i \Omega_i$,
$\Omega_j\cap \Omega_j = \emptyset,\ \forall i\neq j$.
Utilizing the procedure described in Ref.~\cite{latorre2011structure} the original Fokker-Planck equation~(\ref{eq:disc-fp})
is discretized, resulting in a time-inhomogeneous Markov jump process in state
space $S = \{1, \cdots, N\}$, with a time-dependent rate
matrix $\vect L(t) \in \mathbb R^{N\times N}$ satisfying
\begin{align}\label{eq:disc-4}
\sum\limits_{j=1}^N L_{ij}(t) & =  0\\ \label{eq:disc-5}
L_{ij}(t) & \ge  0, \quad i\not= j\\
L_{ij}(t) & =  L_{ij}(t+T)
\end{align}
for all real time $t\geq 0$.
Moreover, the rate matrix $L$ has the form $\vect L(t)=\vect L_0+E(t)\vect L_1$,
where $E(t)$ is periodic, with period $T>0$.
In analogy to~\eqref{eq:disc-fp}, the Markov jump process generated by
$\vect L(t)$ transports probability distributions according to the associated Master equation
\begin{align}
  \label{eq:disc-master}
  \frac{d\vect p(t)}{dt} = \vect L^{\top}(t)\cdot \vect p(t),
\end{align}
where $\vect L^{\top}(t)$ denotes the matrix transpose of $\vect L(t)$. $\vect p(t)$ is an $N$-vector denoting the probability distribution on $S$ at time $t$; $p(i,t)$, for example, is the probability of being in state $i$ (which corresponds to set $\Omega_i$) at time $t$.
As usual,  properties (\ref{eq:disc-4}) and (\ref{eq:disc-5}) of
$\vect L(t)$ guarantee that the total probability mass is conserved,
i.e., if $p(i,0)\ge 0$ component-wise, then $p(i,t)\ge 0$ and $\sum_i
p(i,t) = \sum_ip(i,0)$.
% The solution of the master equation need no be
% periodic.
The temporal evolution of the probability distribution $\vect p(t)$ can be formally written as
\begin{align}  \label{eq:disc-8}
\vect p(t)=\myphi(t)\vect p(0)
\end{align}
by using the
associated propagator matrix $\myphi(t)\in\mathbb R^{N\times N}$ that solves
\begin{align}
  \label{eq:disc-master-phi}
  \frac{d}{dt}\myphi(t) = \vect L^{\top}(t)\myphi(t), \quad \myphi(0) = \id .
\end{align}
Since the last equation can be considered column-wise, the propagator matrix inherits column-wise conservation properties:
$\Phi_{ij}(t) \ge  0$
and $\sum\limits_{i=1}^N \Phi_{ij}(t)  =  1$,
that is, $\myphi^{\top}(t)$ is a
stochastic matrix satisfying $\myphi^{\top}(t)\vect e=\vect e$,
with $\vect e=(1,\ldots,1)^{\top}\in \mathbb R^N$.
Regarding these considerations, we find
\begin{align}
\label{eq:disc-10}  
\myphi_{ij}(t)=\prob\left(\vect X_t=i\mid \vect X_0=j \right),
\end{align}
where $\vect X_t$ denotes the Markov process generated by $\vect L(t)$.

The discretization sets that we used to go from $\vect x_t$ and $\ml(t)$ to $\vect X_t$
and $\vect L(t)$, respectively, can be assumed to provide an arbitrarily fine
partitioning of the original state space; then, the transport properties
of $\vect L(t)$ are almost perfect approximations of the transport properties
of $\ml(t)$, in particular, the approximation $p(i,t)\approx \prob(x_t\in \Omega_i)$ is almost perfect.

@@

As an effect of the periodicity of $\vect L(t)$, the propagator $\myphi(t+T)$
satisfies
\begin{equation}\label{compo-1}
\myphi(t+T)=\myphi(t)\myphi(T),
\end{equation}
for all $t\ge 0$. This can be seen by considering $\vect Y(t)=\myphi(t+T)$. It satisfies
\[
\frac{d\ }{dt}\vect Y(t)=\vect L^{\top}(t+T) \vect Y(t)=\vect L^{\top}(t)\vect Y(t),\quad \vect Y(0)=\myphi(T).
\]
When we consider this identity column-wise and use the propagator property of $\myphi(t)$, we get $\myphi(t+T)=\vect Y(t)=\myphi(t)\myphi(T)$. As a consequence of (\ref{compo-1}), we get for all integers $m=0,1,2,\ldots$ that 
\begin{equation}\label{compo-2}
\myphi(t+mT)=\myphi(t)\myphi^m(T).
\end{equation}
In combination with Eq.~\eqref{eq:disc-8}, we therefore
know the solution $\vect p(t)$ of the Master equation for all $t\ge 0$,
if we can compute $\myphi(t)$ for $t\in (0,T)$.
This is known as the Floquet theorem~\cite{floquet1883equations}.
In particular, we get the long-term evolution of the propagator:
\begin{align}
\label{eq:floq-13}  
\myphi(mT)=\myphi^m(T),
\end{align}
where $\myphi^m(T)$ denotes the $m$th power of $\myphi(T)$. Thus, for the probability at integral periods, we have
\begin{align}
  \label{eq:floq-dynamics}
  p(mT) =  \myphi(mT)\, p(0) = \myphi^m(T)\, p(0).
\end{align}

Using the Floquet theorem, the time-inhomogeneous Markov jump process $X_t$
is therefore discretized into a \emph{time-homogeneous} (not necessarily
reversible) Markov jump process $\tilde X_{m} = X_{mT}, \ m\in\mathbb
N$, which is generated by the transition matrix
\begin{equation}\label{P}
\vect P=\myphi^{\top}(T).
\end{equation}
We prefer to consider the discrete-time process $\tilde X_{m}$ instead of the time-continuous process $X_t$,
because the powerful theories and
computational tools for time-homogeneous Markov processes can be directly applied. It is worth noting that many of these tools require the transition matrix $P$ to satisfy the detailed balance condition. The computations in Appendix~\ref{sec:app-revs} show that $P$ will in general \emph{not} satisfy this condition; in fact,  the deviation {from} reversibility can be estimated from the work of the periodic driving does to the system.
There is no doubt that information within one period is lost by using this temporal
discretization; however, information regarding the long-term behavior of the system on timescales
much longer than the period will be perfectly described, because of $\tilde X_{m} = X_{mT}\approx x_{mT}$ whenever our spatial discretization is sufficiently fine.
At the same time,
the computational cost of generating $\tilde X_{m}$ is much less demanding
than the brute force simulations of NEMD, which implies lower
statistical uncertainty in calculating the observables of interest.

Since $P$ is a stochastic matrix, its eigenvalues are contained in the unit circle in
the complex plane, i.e., each eigenvalue $\lambda$ (potentially complex-valued)
satisfies $|\lambda|\le 1$. Furthermore $\lambda=1$ is an
eigenvalue with a right eigenvector $\vect e=(1,\ldots,1)^{\top}$ and a left eigenvector $\mymu$
satisfying
$\mymu^{\top}P=\mymu^{\top}$.
From now on, we assume $P$ to be irreducible and aperiodic, such that the Perron-Frobenius theorem holds; thus, the eigenvector corresponding to the eigenvalue $\lambda=1$ is  non-negative component-wise, and unique (up to normalization $\sum_i\mu(i)=1$). In this case, $\mymu$ is the stationary measure, in the sense that
$
 \mymu^{\top} P^m = \mymu^{\top},\ m\in\mathbb N
$,
% \begin{align}
% \label{eq:floq-14}  
% \end{align}
and (more precisely) the asymptotic evolution of an initial probability distribution $\vect p(t=0)$ by the process satisfies
$
\vect p^{\top}(0)\,P^m \to \mymu^{\top},\ m\to\infty,
$
% \begin{align}
% \label{eq:floq-15}  
% \end{align}
so that $\mymu$ can be seen as the quasi-stationary distribution of the non-stationary process.

% due to the discretization in both the spacial and temporal directions.
% In the numercal example in Section~\ref{sec:alanine}, we
% compute the

@@

% \subsection{Building the Markov State Model from the Floquet transition matrix}
If the discretization cells $\Omega_i$, $i=1,\ldots,N$, form a fine
partition of the molecular state space, the Markov chain defined via
the transition matrix $P$
 is discrete in time, but in space it remains a fine-scale description of the transport properties of the
dynamics with a very large number ($N$) of states.  Now we want to coarse our description much further by
constructing a Markov State Model (MSM) for $\vect P$ with $K\ll N$
\emph{macrostates}, which would be the metastable states of the system. The resulting $K\times K$ MSM transition matrix $\hat{\vect P}$
then defines the coarse-grained long-term kinetics that shall
approximate the original long term kinetics well. 
The idea behind MSM building is that, given that the molecular system under consideration exhibits metastable conformations,  it is usually possible to construct a
relatively small number of discrete sets --the metastable sets that form the so-called macrostates-- that
correctly describe the slow  dynamics, and 
in each set the fast dynamics relaxes on some timescales that are significantly shorter than the metastable timescales.
Then, if the MSM dynamics reproduces the slow timescales and the corresponding transitions
of the original dynamics~\eqref{eq:disc-1},
the former is considered to be a good approximation of the latter.

MSM building has attracted much attention recently, and extensive theories~\cite{A19-31} as well as algorithms \cite{A19-1}, applications (see e.g. \cite{A19-26,PNAS09} for two of hundreds of articles), and software \cite{A19-49, MSMBuilder} have been developed. However, 
by far most of the literature is related to building \emph{standard} MSMs for equilibrium MD. In standard MSM, the transition region also has to be discretized, a feature that often forces the user to incorporate more macrostates than essentially needed to approximate the long-term kinetics.
In Ref.~\cite{sarich2010approximation, A19-31,schuette2011markov,BucheteHummer} it is shown how to construct a \emph{non-standard} MSM in a way that avoids this problem for equilibrium MD, i.e., if $\vect P$ satisfies the detailed balance
condition: (1) identify the cores of the metastable sets of the
dynamics, (2) use them as milestones to construct an MSM in which the
macrostates are the metastable core sets and $\hat{\vect P}$ is the transition
matrix of the milestone process \cite{A19-31,schuette2011markov,A19-29} that models the jumping behavior of
the original dynamics between the metastable regions.
